---
title: ' Data Profiling with Pandas in R'
author: Brett R. Taylor
date: '2020-08-17'
slug: data-profiling-with-pandas-in-r
categories:
  - Data
  - Science
tags:
  - Data
  - Database
  - R
  - Python
output:
  html_document:
    df_print: paged
    highlight: zenburn
    theme: journal
---

# Data Profiling

Profiling data in SQL is possible.  There are many other tools today that are better at analyzing the data, and this will increase in the future.  

```{r init, echo=TRUE, message=FALSE, warning=FALSE}
library(DBI)
library(reticulate)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(progress = FALSE,verbose = FALSE)
use_condaenv('mdata',conda = '/opt/conda/bin/conda',required = TRUE)
 

```







```{python}
import sys
from  pandas import read_csv,concat
 
df1 =  read_csv("/data_source/public_data/phi_data/breach_report_2020_08_06_archive.csv")
df2 =  read_csv("/data_source/public_data/phi_data/breach_report_2020_08_06_investigate.csv")
df =  concat([df1,df2],ignore_index=True) 
```


 

 
## Profiling DataFrames using Data Science Tools

Data Science has evolved over the last thirteen years, and one of the best tools for understanding tables in databases, and CSV file sources is based on Python.  The Python tool that should be used is pandas-profiling which is opensource, and avaiable on [GitHub](https://github.com/pandas-profiling/pandas-profiling)

This is an example of how to display the analysis of pandas-profiling in the markdown file.

```{python echo=TRUE, error=TRUE, message=TRUE}

import sys
import pandas as pd
import pandas_profiling as pdp
import matplotlib
 
df1 = pd.read_csv("/data_source/public_data/phi_data/breach_report_2020_08_06_archive.csv")
df2 = pd.read_csv("/data_source/public_data/phi_data/breach_report_2020_08_06_investigate.csv")
df = pd.concat([df1,df2],ignore_index=True) 
profile = df.profile_report(title='Portfolio Report', plot={'histogram': {'bins': 8}},progress_bar=True)

```
 
 
## Error Occurs and displaying HTML on Markdown fails 
 
```{r display_pandas_profile_in_rmarkdown, echo=TRUE, error=TRUE, message=FALSE}
library(shiny)

 tags$div(
                HTML(py$profile$to_html()  )
 )  
 
```


# Fixing the RMarkdown and Shiny problem

There is a simple solution to make the Python based Pandas-Profiling package to display in R Markdown systems, and the Shiny app.   During investigation of the problem, we identified that you must use "Agg" which is a ``non-interactive backend`` thread.


Displaying the profile using shiny works in the Rmarkdown environment.

```{r display_pandas_profile_in_rmarkdown_works, echo=TRUE, error=FALSE, message=FALSE}
library(shiny)
reticulate::import("matplotlib") 
py$matplotlib$use("Agg")
tags$iframe(srcdoc=HTML(py$profile$to_html()), height=800, width=900)
 
```

```{python eval=FALSE, include=FALSE}

# Load Packages
import numpy as np
import pandas as pd

import matplotlib as mpl
import matplotlib.pyplot as plt
mpl.use("Agg")

# import sklearn
# from sklearn.datasets import load_boston

from pandas_profiling import ProfileReport

# Get Dataset
#boston = load_boston()

boston_df = pd.read_csv("/data_source/public_data/phi_data/breach_report_2020_08_06_investigate.csv")

#boston_df = pd.DataFrame(boston.data, columns = boston.feature_names)

# Make Profiling Report
boston_profile = ProfileReport(boston_df, title = "Boston Profile Report")

boston_profile.to_file("boston_profile.html") # <-- FAILURE HAPPENS HERE

```

